### 2024-04-29 paper.tex
minor
shorten paper

\subsubsection{Correlating VMR methylation with gene expression}

@@

\subsubsection{Improved identification of cell types}

@@

\subsubsection{Robustness to parameter changes}

@@

\subsubsection{Further applications}

@@

We have implemented the approach just described in a Python package with a command line interface, \textit{MethSCAn} (ANalysis of Single-Cell METHylation data), which also offers a number of other functionalities for the analysis of scBS data (Fig.~\ref{fig:workflow}).
The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or BISCUIT \citep{biscuit}.
For each cell, these tools produce a text file that lists the methylation status of all CpG sites.
Since it is inconvenient to work with hundreds or thousands of text files, \textit{MethSCAn} provides the \texttt{prepare} command which parses these methylation files and stores their content in a compressed format that enables efficient access to all CpG sites of the genome.
\texttt{methscan prepare} is flexible and accepts all tabular input formats including Bismark .cov-files, methylpy .allc-files, or custom user-defined formats.

In brief, we store methylation of each chromosome in a matrix where each column represents a cell and each row represents a base pair.
Methylated and unmethylated sites are coded as $+1$ and $-1$, respectively.
Since the vast majority of values in this matrix are missing due to the sparsity of scBS data (and because rows for base-pairs not corresponding to a CpG site contain no data), we encode missing values as zero and then store the data in Compressed Sparse Row (CSR) format.
This format does not explicitly store zeroes (here: missing values) and is optimized for row-wise access, which results in significant compression and allows fast access to the methylation status of genomic intervals.

\texttt{methscan prepare} also computes a number of summary statistics for each cell, including the mean genome-wide methylation level and the number of observed CpG sites, i.e.\ the number of CpG sites that have sequencing coverage.
These summary statistics can be used to detect cells with poor quality.
The quality of each single cell methylome can furthermore be inspected with \texttt{methscan profile}, which computes the average methylation profile of a set of user-defined genomic regions such as transcription start sites (TSS) at single-cell resolution.
The TSS profile is a useful quality control plot since methylation shows a characteristic dip roughly \textpm1~kb around the transcription start site in mammalian genomes.
Cells that do not show this pattern, or cells with few observed CpG sites, may then be discarded with \texttt{methscan filter}.

After quality control, the user has access to the genome-wide VMR detection approach described earlier, using the \texttt{methscan scan} command.
This produces a BED file that lists the genome coordinates of VMRs.
To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED file can be used as input for \texttt{methscan matrix}, which quantifies methylation at genomic intervals in all cells.
The command produces both, the simple percentage of methylated CpG sites, as well as our proposed methylation measure, the shrunken mean of the residuals, which is more robust to variation in sequencing coverage or stochastic differences in read position between cells.
We note that both \texttt{matrix} and \texttt{profile} accept any valid BED file as input, which means that the user can quantify and visualize methylation at any set of genomic features of interest, such as promoters, enhancers or transcription factor binding sites, obtaining one profile plot per cell.
The methylation matrix produced by the \texttt{matrix} function can then serve as input for established methods used for the exploration of single-cell data, such as dimensionality reduction and cell clustering.

Finally, after annotation and exploration of the data set, the user may specify two groups of cells for DMR detection with \texttt{methscan diff}.
This command produces a BED file listing the genome coordinates, adjusted p-values, as well as several other metrics associated with DMRs.
These DMRs may then be associated with nearby genes, or used for GO enrichment with tools such as GREAT \citep{mclean2010great} to enable functional interpretation.

@@

\subsection*{Author Contributions}
MMB developed and implemented DMR detection under supervision of LPMK and SA.
SO, LK and LPMK performed benchmarks.
LPMK and SA wrote the manuscript.
% TO DO: complete this

@@

\begin{figure*}[p]
	\begin{center}
		\includegraphics[width=.8\textwidth]{figures/Fig_workflow.png}
	\end{center}
	\caption{\small \textbf{Overview of the functionalities implemented in the \textit{MethSCAn} package.}\\
		\texttt{methscan prepare} parses methylation files produced by common bisulfite sequencing mappers and stores their contents in a compressed format optimized for efficient access to genomic intervals.
		\textit{MethSCAn} also produces cell-wise summary statistics and quality plots (here: average methylation around the transcription start site) that are used to detect low-quality cells.
		These cells can be discarded with \texttt{methscan filter}.
		To obtain a methylation matrix, similar to the count matrices used in scRNA-seq, the user must first decide in which genomic intervals methylation should be quantified.
		The user may either provide genomic regions of \emph{a priori} interest, or they may choose to discover VMRs (variably methylated regions) in the data with \texttt{methscan scan}.
		The resulting methylation matrix can then be used for downstream analysis such as cell clustering and dimensionality reduction.
		Identified groups of cells (e.g. clusters) can then be scanned for DMRs with \texttt{methscan diff}.
	}
	\label{fig:workflow}
\end{figure*}

@@

The function \texttt{methscan prepare} reads a set of methylation files (e.g. produced by Bismark) and produces a file that stores the matrix $x$ in a space-efficient format, as follows:
$x$ is represented as a SciPy sparse matrix \citep{SciPy}, encoding the actual values 0, 1, and NA as -1, 1, and 0, respectively.
Coding NA (the most common value) as 0 leverages SciPy's sparse matrix storage.
In all that follows here, any mention of $x$ will, however, always mean the encoding as $x_{ij}\in\{0,1,\text{NA}\}$.

>>>>

\subsubsection*{Correlating VMR methylation with gene expression}

@@

\subsubsection*{Improved identification of cell types}

@@

\subsubsection*{Robustness to parameter changes}

@@

\subsubsection*{Further applications}

@@

We have implemented the methods just described in a Python package with a command line interface, \textit{MethSCAn} (Single-Cell ANalysis of METHylation data), which also offers a number of other functionalities for the analysis of scBS data. 
Fig.~\ref{fig:workflow} illustrates a typical scBS data analysis with \texttt{MethSCAn} and provides an overview of the implemented core functionalities.
The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or BISCUIT \citep{biscuit}, which are first stored in a compressed format that enables efficient access to all CpG sites of the genome (see Methods).

@@

\subsection*{Author Contributions}

MMB developed and implemented DMR detection under supervision of LPMK and SA.
SO, LK and LPMK performed benchmarks.
LPMK and SA wrote the manuscript.
% TO DO: complete this

@@

\begin{figure*}[p]
	\begin{center}
		\includegraphics[width=.8\textwidth]{figures/Fig_workflow.png}
	\end{center}
	\caption{\small \textbf{Overview of the functionalities implemented in the \textit{MethSCAn} package.}\\
		Common bisulfite sequencing mappers produce one methylation report file per cell.
		As it is inconvenient to work with thousands of text files, \texttt{methscan prepare} first stores their contents in a compressed format optimized for efficient access to genomic intervals.
		This command also produces cell-wise summary statistics that can be used to discard low-quality cells with \texttt{methscan filter}.
		The quality of each cell can furthermore be assessed by visualizing the average methylation around transcription start sites, which are known to show a characteristic dip in mammalian genomes, using \texttt{methscan profile}.
		This command can also visualize methylation at other genomic features of interest, such as enhancers or transcription factor binding sites.
		To obtain a methylation matrix analogous to scRNA-seq count matrices, the user must first decide in which genomic intervals methylation should be quantified.
		The user may either provide genomic regions of \emph{a priori} interest as a BED file, or they may choose to discover VMRs (variably methylated regions) in the data with \texttt{methscan scan}.
		The \texttt{methscan matrix} command then quantifies methylation at these genomic intervals in all cells.
		It provides two metrics:
		the simple percentage of methylated CpG sites and our proposed measure, the shrunken mean of the residuals, which is more robust to variation in sequencing coverage and stochastic differences in read position between cells.
		The resulting methylation matrix can then serve as input for established methods used for the exploration of single-cell data, such as dimensionality reduction and cell clustering.
		Finally, after annotation and exploration of the data set, the user may specify two groups of cells for DMR (differentially methylated region) detection with \texttt{methscan diff}.
		This command produces a BED file listing the genome coordinates, adjusted p-values, as well as several other metrics associated with DMRs.
		These DMRs may then be associated with nearby genes, or used for GO enrichment with tools such as GREAT \citep{mclean2010great} to enable functional interpretation.
	}
	\label{fig:workflow}
\end{figure*}

@@

The function \texttt{methscan prepare} reads a set of methylation files (e.g. produced by Bismark) and produces one file per chromosome.
These files store the matrix $x$, where each column represents a cell and each row represents a base pair, in a space-efficient format as follows:
$x$ is represented as a SciPy sparse matrix \citep{SciPy}, encoding the actual values 0, 1, and NA as -1, 1, and 0, respectively.
Since the vast majority of values in this matrix are missing due to the sparsity of scBS data (and because rows for base-pairs not corresponding to a CpG site contain no data), we encode missing values as zero and then store the data in Compressed Sparse Row (CSR) format.
This format does not explicitly store zeroes (here: missing values) and is optimized for row-wise access, which results in significant compression and allows fast access to the methylation status of genomic intervals.
In all that follows here, any mention of $x$ will, however, always mean the encoding as $x_{ij}\in\{0,1,\text{NA}\}$.
